---
title: "Coursera Reproducible Research week2 assinment"
author: "Evangelos Zormpas"
date: "July 6, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This assignment involves exploratory data analysis on data from a personal activity monitoring device, and presentation of analysis and results using R markdown.

#### First load required Libraries
```{r message=FALSE}
library (dplyr)
library (ggplot2)
```

### 1. Code for reading in the dataset and/or processing the data
Unzip and then read the datafile using read.csv.
```{r}
unzip ("activity.zip")
a <- read.csv("activity.csv")
```

Create a copy of data ommiting NA Values.
```{r}
a1 <- na.omit(a)
```

### 2. Histogram of the total number of steps taken each day
I use dplyr functions to group by and summarize data by date and ggplot2 package to draw histogram.  

```{r}
a1_date <-group_by(a1, date) 
a1_sum_date <- summarise(a1_date, steps_sum = sum( steps))
qplot(steps_sum,  data=a1_sum_date)
```

### 3. Mean and median number of steps taken each day
Calling summary on summarized data per date will return mean and median (among others)  
```{r}
summary(a1_sum_date)
```

### 4. Time series plot of the average number of steps taken
We will have to group by the 5 min intervals and summarize on steps and then plot sum of steps by intervals. 
```{r}
a1_5min <-group_by(a1, interval) 
a1_sum_5min <- summarise(a1_5min, steps_sum = sum(steps))
ggplot(data=a1_sum_5min, aes(x=interval, y=steps_sum)) + geom_line()
```


### 5. The 5-minute interval that, on average, contains the maximum number of steps

Get the row containing max steps (by 5min interval) and then select interval column. 
```{r}
a1_sum_5min[which.max(a1_sum_5min$steps_sum),]$interval
```

### 6. Code to describe and show a strategy for imputing missing data

There are `r sum(is.na(a$steps))` missing values from the dataset. 
(the number of missing values is produced by the r code  "sum(is.na(a$steps))")

Each NA in the steps valriable of the original dataset will be replaced by the mean (of all days) of this 5 minute interval. 

* First calculate mean of all days for each interval. 
```{r}
a1_avg_5min <- summarise(a1_5min, steps_mean = mean(steps))
```

* Then merge original data with mean data on interval 
```{r}
a_mean <- merge (a, a1_avg_5min, by="interval")
```

* Then Create a column that when steps is NA replaces it with mean. 
And then leave out the speps_mean column
```{r}
a_mean <- mutate (a_mean, steps=ifelse(is.na(steps),steps_mean,steps ))
a_mean <- select (a_mean, interval:date)
```

### 7.Histogram of the total number of steps taken each day after missing values are imputed
```{r}
amean_date <-group_by(a_mean, date) 
amean_sum_date <- summarise(amean_date, steps_sum = sum(steps))
qplot(steps_sum,  data=amean_sum_date)
```

Calling summary on summarized data per date will return mean and median (among others)  
```{r}
summary(amean_sum_date)
```

Comparing to the mean and median of the original data, there is only a small change in median. Imputed dataset has median  10.766 instead of 10.765. 

### 8.Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

We are going to use the imputed dataset. 
First create a variable day_type separating weekdays from weekends:
```{r}
a_mean1 <- mutate (a_mean, day_type=ifelse(weekdays(as.Date(a_mean$date)) %in% c("Saturday","Sunday"),"weekends","weekdays" ))
```

Then we group by interval and day_type  and plot using ggplot and facets
```{r}
amean1_sum_5min <-a_mean1 %>% group_by( interval, day_type) %>%  summarise(steps_avg = mean(steps))
ggplot(amean1_sum_5min, aes(x=interval, y=steps_avg)) + geom_line() + facet_grid(day_type ~ .)
```


